#Reproducible Research: Peer Assessment 1

###John Robinson###

####Loading and preprocessing the data

The data file activity.csv is loaded and the interval column set as a factor in variable activityData

```{r Data loading and preparation}
setwd("C:/Users/John/git/RepData1") # Set the working directory
unzip("activity.zip") # Unzip the data
activityData <- read.csv("activity.csv") # Load the data into a variable
activityData$interval <- as.factor(activityData$interval)
```

####What is mean total number of steps taken per day?

The average number of steps per day is shown in the histogram

```{r Histogram and summary of base data Steps per day, fig.height=4,fig.width=4}

hist(as.numeric(tapply(activityData$steps,activityData$date,sum)),main="Steps per day", xlab="Steps per day",breaks=61,col="Red")

summary(tapply(activityData$steps,activityData$date,sum))


```

The data show a mean number of steps of 10770 with a median value of 10760  


#### What is the average daily activity pattern?

This is shown in the following plot of the average steps per five minute interval, ignoring any rows with NA values  

```{r Averaged steps per five minutes, fig.height=4,fig.width=4}
meanSteps <- tapply(activityData$steps,activityData$interval,mean,na.rm=TRUE) # Create a variable of the average steps per period
plot(meanSteps,xlab="Five minute period",ylab="Average steps taken",col="blue",type="l") # Time series plot of the means

```

The five minute period which contains the maximum number of average steps can be determined by finding the point at which the maximum occurs in the variable meanSteps as follows:

```{r Interval containing the maximum}
names(meanSteps[meanSteps[]==max(meanSteps)])
```

This shows that the maximum mean activity appears to occur at 8:35 in the morning

#### Imputing missing values

Examination of the original data shows that there are `r sum(is.na(activityData))` missing values in the data

In order to address this, the missing values have been replaced by the average number of steps in the equivalent five minute period, taken over all the days of the trial, in a new variable called fullactivityData

```{r Replacing missing data}
fullactivityData <- activityData  # Create variable
for (i in 1:dim(fullactivityData)[1]){  #loop through all elements
     if (is.na(fullactivityData[i,1])) { #if the element is NA replace with the mean previously calculaed in meanSteps
            fullactivityData[i,1] <- as.integer(meanSteps[names(meanSteps)==as.character(fullactivityData[i,3])])
     }   
}
```

The revised average number of steps per day is shown in the histogram

```{r Revised histogram and summary of base data Steps per day, fig.height=4,fig.width=4}

hist(as.numeric(tapply(fullactivityData$steps,fullactivityData$date,sum)),main="Steps per day", xlab="Steps per day",breaks=61,col="Red")

summary(tapply(fullactivityData$steps,fullactivityData$date,sum))


```

The revised data show a mean number of steps of 10750 with a median value of 10640, as compared with the previos average of 10770 and median of 10760.

The addition of the estimates, taken from the calculated mean steps per five minute intervals, has left the average close to its previous value, but shifted the median to the left of the dstribution.


#### Are there differences in activity patterns between weekdays and weekends?

To investigate differences in activity between week days and weekends, a new column was created of the day names, called days.This column was then parsed to a logical vector, to identify weekend days, and turned into a factor.

```{r Create a factor vriable for weekends}
fullactivityData <- cbind(fullactivityData,fullactivityData$date) # Create an extra column
names(fullactivityData)[4]="days" # Rename the column
fullactivityData$days <- weekdays(as.Date(fullactivityData$days)) # Create a vector of day names
fullactivityData$days <- fullactivityData$days[] %in% c("Saturday","Sunday") # Parse for weekends
fullactivityData$days <- factor(fullactivityData$days,labels=c("Weekday","Weekend")) # Turn it into a factor
```

The full set of activity data is then interrograted to determine the mean number of steps on weekdays and weekends, and plot a smoothed view of the data.

```{r Determine weekday and weekend means}
weekendActivity <- fullactivityData[fullactivityData$days=="Weekend",] 
weekdayActivity <- fullactivityData[fullactivityData$days=="Weekday",]
weekendSteps <- tapply(weekendActivity$steps,weekendActivity$interval,mean) # Create a variable of the average steps per period
weekdaySteps <- tapply(weekdayActivity$steps,weekdayActivity$interval,mean)
stepData <- data.frame(c(as.numeric(weekendSteps),as.numeric(weekdaySteps)),c(1:288,1:288),c(rep("Weekend",times=288),rep("Weekday",times=288))) # Create a data frame for plotting results
names(stepData) <- c("Steps","Interval","Days")                      
library("lattice", lib.loc="D:/Programs/R/R-3.1.1/library")
xyplot(Steps~Interval|Days,data=stepData,type=c("spline","g"),layout=c(1,2))
```

The differences beetween weekday and weekend activity are clear from the plot. Weekend activity is more evenly spread over the day, with a lower morning peak.
